A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-03-19 lctes18-reviews-11-rebuttal.txt

We would like to thank the reviewers for the thoughtful comments, questions,
and criticism.

In a final version, we will address the main concern among reviews regarding
the paper motivations.
We will discuss why having a formal semantics of the language is important, and
how it affected our own implementation.
We will also discuss why the main results of the paper (memory boundedness,
determinism, and termination) are particularly important in the context of
embedded systems.

We will accommodate all requested clarifications and corrections in a final
version of the paper (more intuitive/complete descriptions of the semantics,
compare with some specific work, etc.)
Some of the comments are also addressed individually as follows.

Review #11A
===========

> Concerning the language itself, it is clear indeed that the strong
> restrictions <...> help to ensure the properties which are proved.
> However, even if it is the case for other languages <...>, the fact of
> linking the semantics of the language to what could be a particular
> implementation (scheduling in lexical order) could be considered somewhat
> questionable. While the language uses, in principle, parallel constructions,
> the parallelism seems in any case very largely limited by such a choice.

Considering the constant and low-level interactions with the underlying
architecture in embedded systems (e.g., direct port manipulation), we believe
that lexical scheduling must be part of the language specification as a
pragmatic design decision.

However, Céu has compile-time mechanisms to detect trails in parallel that
never share resources (`[Section 2.2]`).
In such cases, an implementation could take advantage of real parallelism.

Review #11B
===========

> - There are no semantics given for certain language features, e.g. what
>   happens when a call to a C function is made using the underscore syntax?
>   Also, it’s mentioned that side effects and system calls behave like in
>   conventional imperative languages, and it would be good to have a reference
>   for what this behaviour is.

In this context, C calls and system calls are synonymous.
The behavior for calls and side effects, all encapsulated in the `mem(id)`
primitive, is to modify a memory location in "instantaneous" time.
They are opaque operations that have no effect on the application execution
flow (e.g., they cannot block or abort execution).

> Are there approaches to this distributed synchronous programming problem
> other than the event-driven model?

Most synchronous languages have a clock/tick-based model of time
(`[Section 4]`).
These models are duals since we can accommodate a single event in a single tick
and vice versa.

Review #11C
===========

> How does the focus on the control aspects of the language (and leaving out
> side-effects and system calls) affect the semantics and the conclusions which
> can be drawn from analysing it?

The focus on control does not affect the logical semantics, abstract program
behavior (independent of real-time considerations).
Any modeling of side effects would have to assume, or somehow guarantee, not to
break the synchronous model (e.g., that system calls are non-blocking).
Such assumptions or guarantees would lead us to the same result.

Review #11D
===========

> How is determinism ensured across different runs for the example in fig 2a,
> if events A and B can occur in any order?

The program behaves deterministically in the sense that the same inputs (event
history) will necessarily produce the same output.

> Overall, the paper gives the impression of proposing CEU, <...>. Is the
> contribution the actual formalization of these semantics?

Yes, the contribution is the actual formalization and proofs of memory
boundedness, determinism, and termination.

> Otherwise, how was synchronization between traits implemented previously in
> CEU to ensure correctness?

From the very beginning, the language has been designed (and implemented) with
these safety properties in mind.
The proposed formalization of this paper has helped us to validate the design
and has also influenced the current implementation.

> Can await-emit lead to a deadlock if called from two concurrent trails,
> waiting on each other?

No.
An `emit` remains pending until its triggered internal reaction completes.
If it is pending it cannot be in an `await`, so a new `emit` in the triggered
reaction cannot retrigger the original `emit`.
Stack-based internal events prevents code reentrancy that would occur with a
queue-based policy.

> How does the compiler ensure that no non-blocking operations are called from
> CEU? Is there a black list of library calls?

The actual implementation of Céu supports a command-line option that accepts a
white list of library calls.
If a program tries to use a function not in this list, the compiler raises an
error.

### 2018-03-19 lctes18-reviews-11-rebuttal.txt

We would like to thank the reviewers for the thoughtful comments, questions,
and criticism.

In a final version, we will address the main concern among reviews regarding
the paper motivations.
We will discuss why having a formal semantics of the language is important, and
how it affected our own implementation.
We will also discuss why the main results of the paper (memory boundedness,
determinism, and termination) are particularly important in the context of
embedded systems.

We will also accommodate all requested clarifications and corrections (e.g.,
more intuitive/complete descriptions of the semantics, compare with some
specific work, etc.)
Some of the comments are also addressed individually as follows.

@@

The focus on control does not affect the logical semantics, abstract program
behavior (independent of real-time considerations).
Any modeling of side effects would have to assume, or somehow guarantee, not to
break the synchronous model (e.g., that system calls are non-blocking).
Such assumptions or guarantees would lead us to the same result.

Review #11D
===========

> How is determinism ensured across different runs for the example in fig 2a,
> if events A and B can occur in any order?

The program behaves deterministically in the sense that the same inputs (event
history) will necessarily produce the same output.
So, a history where `A` occurs before `B` is different from a history where `B`
occurs first.
Different inputs can produce different outputs.

> It is not clear how many of these semantics exist in the implementation of
> CEU presented in TECS'17. <...>.
> Overall, the paper gives the impression of proposing CEU, <...>. Is the
> contribution the actual formalization of these semantics?

Yes, the contribution is the actual formalization and proofs of memory
boundedness, determinism, and termination.
Céu itself has already been presented in other publications such as TECS'17.

### 2018-03-22 01-intro.tex
Revisões do artigo.

\CEU inherits the synchronous and imperative mindset of Esterel but adopts a
simpler semantics with fine-grained execution control~\cite{ceu.tecs17}.
%
The list that follows summarizes the semantic peculiarities of \CEU:
%
\begin{itemize}
    \item Fine-grained, intra-reaction deterministic execution, which makes
          \CEU fully deterministic.
    \item Stack-based execution for internal events, which provides a limited
          but memory-bounded form of subroutines.
    \item Finalization mechanism for lines of execution, which makes abortion
          safe with regard to external resources.
\end{itemize}

In this work, we present a formal small-step structural operational
semantics for \CEU and prove that it enforces memory-bounded, deterministic,
and terminating reactions to the environment, i.e., that for a given
arbitrary timeline of input events, multiple executions of the same program
always react in bounded time and arrive at the same final finite memory
state.

% TODO: \fs{Descrever seções.}

>>>>

\CEU inherits the synchronous and imperative mindset of Esterel but adopts a
simpler semantics with fine-grained execution control.
%
The list that follows summarizes the semantic peculiarities of
\CEU~\cite{ceu.tecs17}:
%
\begin{itemize}
    \item Fine-grained, intra-reaction deterministic execution, which makes
          \CEU fully deterministic.
    \item Stack-based execution for internal events, which provides a limited
          but memory-bounded form of subroutines.
    \item Finalization mechanism for lines of execution, which makes abortion
          safe with regard to external resources.
\end{itemize}

In this work, we present a formal small-step structural operational
semantics for \CEU and prove that it enforces memory-bounded, deterministic,
and terminating reactions to the environment, i.e., that for a given
arbitrary timeline of input events, multiple executions of the same program
always react in bounded time and arrive at the same final finite memory
state.
%
Conceiving a formal semantics for \CEU leads to a number of capabilities and
outcomes as follows:

\begin{enumerate}
\item
    Understanding, finding corner cases, and setting compromises.
    After the semantics was complete and discussed in extent, we found a bug in
    the order of execution of statements in a specific case.
\item
    Explaining core aspects of the language in a reduced, precise, and
    unambiguous way.
    This is particularly important when comparing languages that are similar in
    the surface (e.g., \CEU and Esterel).
\item
    Implementing the language.
    Small-step operational semantics typically describe abstract machines that
    are close to concrete implementations.
    The current implementation of the \CEU scheduler is based on the formal
    semantics.
\item
    Proving properties for particular or all programs in the language.
    We prove that all programs in \CEU are memory bounded, deterministic, and
    react in finite time.
\end{enumerate}

The last item is particularly important in the context of embedded systems:

\begin{description}
\item[Memory Boundedness:]
At compile time, we can ensure that the program fits in the device's restricted
memory and that it will not grow unexpectedly during runtime.
\item[Deterministic Execution:]
We can simulate an entire program execution providing an input history with the
guarantee that it will always behave the same.
This can be done in negligible time in a controlled development environment and
with multiple users (e.g., multiple developers each on its desktop) before
deploying the application in the actual device.
\item[Terminating Reactions:]
Real-time applications must guarantee responses within specified deadlines.
A terminating semantics enforces upper bounds for all reactions and guarantees
that programs always progress with the time.
In addition, it also ensures that programs always reach an idle state amenable
to entering in standby mode, which is can improve the autonomy of
battery-powered devices.
\end{description}

### 2018-03-22 03-semantics.tex
Revisões do artigo.

In this section, we introduce a reduced syntax for \CEU and propose an
operational semantics to formally describe the behavior of its programs.
We describe a small synchronous kernel highlighting the peculiarities of \CEU,
in particular, the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language,
leaving out side-effects and system calls (which behave like in conventional
imperative languages).

>>>>

In this section, we introduce a reduced syntax for \CEU and propose an
operational semantics to formally describe the behavior of its programs.
We describe a small synchronous kernel highlighting the peculiarities of \CEU,
in particular, the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language,
and encapsulate all side effects and system calls into a single abstract memory
operation.
Memory operations are opaque and behave like in conventional imperative
languages (i.e., reads and writes), having no effect on the application
execution flow (e.g., they cannot block or abort execution).
